---
title: "Introduction to treetable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to treetable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(treetable)
```

treetable aggregates a flat data frame into an arbitrary-depth hierarchy with
subtotals at every level ([hierarchical_table()]), and renders that hierarchy
as an interactive, expandable/collapsible tree table built on `htmlwidgets`
([tree_table()]). Everything — row colors, fonts, column widths, number
formats and the per-leaf detail modal — is configured from R, without
touching CSS or JavaScript.

## 1. Aggregating a hierarchy with `hierarchical_table()`

`hierarchical_table()` takes a flat data frame, a vector of hierarchy columns
(from most general to most specific), and a vector of numeric columns to sum
at every level:

```{r}
sales <- tibble::tibble(
  area     = c("Sales", "Sales", "Sales", "Costs", "Costs"),
  category = c("Product A", "Product A", "Product B", "Payroll", "Rent"),
  region   = c("North", "South", "North", "North", "South"),
  amount   = c(1000, 750, 500, -1200, -300)
)

hierarchical_table(sales, levels = c("area", "category"), values = "amount")
```

The result is a long-format tibble: one row per node of the tree (`level_1`,
..., `leaf`), with a subtotal already computed for every group and subgroup.
Add as many levels as you need — `hierarchical_table()` recurses through
however many columns you list in `levels`.

## 2. `tree_table()`, reactable-style columns

`tree_table()` renders the same kind of hierarchy as an interactive tree
table. Unlike `hierarchical_table()`, its `columns` is a list of
[col_spec()] groups — each with a header `name` and one or more
[col_field()]s underneath it, every field with its own source column in
`data`, its own aggregation, and its own display format. There's no shared
naming pattern required between a group's display name and its source
column(s); `tree_table()` also expects an `id` column identifying each leaf
row:

```{r}
budget_vs_actual <- tibble::tibble(
  area = c("Sales", "Sales", "Costs"),
  category = c("Product A", "Product B", "Payroll"),
  id = c("v-a", "v-b", "c-payroll"),
  Q1_budget = c(1000, 500, 300), Q1_actual = c(950, 600, 280),
  Q2_budget = c(1100, 550, 300), Q2_actual = c(1200, 500, 310)
)

budget_columns <- list(
  col_spec(
    "Q1",
    col_field("Q1_budget", label = "Budget", format = column_format(type = "currency", decimals = 0)),
    col_field("Q1_actual", label = "Actual", format = column_format(type = "currency", decimals = 0))
  ),
  col_spec(
    "Q2",
    col_field("Q2_budget", label = "Budget", format = column_format(type = "currency", decimals = 0)),
    col_field("Q2_actual", label = "Actual", format = column_format(type = "currency", decimals = 0))
  )
)

tree_table(budget_vs_actual, levels = c("area", "category"), columns = budget_columns)
```

A group with a single field (`col_spec("Q1", col_field("Q1_budget"))`, say)
renders as one plain column. With more than one field, every field just
renders as its own plain column next to the others under the shared group
header — treetable does not compute a comparison between them. If you want a
variance/percentage column, compute it yourself and add it as another
[col_field()], so you decide what it means and how (or whether) it's styled.

## 3. Custom theme, and fully custom cells

Use [treetable_theme()] to restyle the table's colors, fonts and column
widths. `format` (a [column_format()]: currency, number/decimal, percentage,
or text) covers most display needs, but a [col_field()]'s `cell` goes
further, reactable-style: a `function(value, row)` returning an `htmltools`
tag for full control over a cell's HTML — colors, badges, icons, anything —
or `NULL` to fall back to `format` for that row:

```{r}
tree_table(
  budget_vs_actual,
  levels = c("area", "category"),
  columns = list(
    col_spec(
      "Q1",
      col_field(
        "Q1_budget", label = "Budget",
        format = column_format(type = "currency", decimals = 0)
      ),
      col_field(
        "Q1_actual", label = "Actual",
        format = column_format(type = "currency", decimals = 0),
        cell = function(value, row) {
          if (is.na(value)) return(NULL)
          over_1000 <- value > 1000
          htmltools::tags$span(
            style = if (over_1000) "color:#1a7f37;font-weight:600" else NULL,
            paste0("$", format(value, big.mark = ","))
          )
        }
      )
    )
  ),
  theme = treetable_theme(header_bg = "#102a43", subtotal_bg = "#e3f2fd")
)
```

## 4. Per-leaf detail modal in Shiny

Passing a non-`NULL` `detail` to `tree_table()` shows a detail button on
every leaf row. Clicking it sets the Shiny input
`input[[paste0(output_id, "_detail")]]` to `list(id, nonce)`; call
[show_detail_modal()] from an `observeEvent()` to react to it.

### Automatic table

[detail_table()] builds a modal that filters raw, pre-aggregation rows by
leaf id and shows them as a table (using `reactable` when installed, falling
back to `DT`, then to a plain HTML table):

```{r, eval = FALSE}
library(shiny)

raw_movements <- tibble::tibble(
  id = c("v-a", "v-a", "v-b", "c-payroll"),
  date = as.Date(c("2025-01-05", "2025-01-19", "2025-01-11", "2025-01-31")),
  amount = c(600, 400, 500, -300)
)
detail <- detail_table(raw_movements, id_col = "id")

ui <- fluidPage(treetableOutput("tbl"))
server <- function(input, output) {
  output$tbl <- renderTreetable(
    tree_table(
      budget_vs_actual,
      levels = c("area", "category"),
      columns = budget_columns,
      detail = detail
    )
  )
  observeEvent(input$tbl_detail, {
    show_detail_modal(input$tbl_detail$id, raw_movements, detail)
  })
}
shinyApp(ui, server)
```

### Custom UI

Pass a `function(id, raw_data)` instead to render anything you want in the
modal — your own table, a plot, free text:

```{r, eval = FALSE}
custom_detail <- function(id, raw_data) {
  rows <- raw_data[raw_data$id == id, ]
  tagList(
    h4(paste("Movements for", id)),
    renderTable(rows)
  )
}

server <- function(input, output) {
  output$tbl <- renderTreetable(
    tree_table(
      budget_vs_actual,
      levels = c("area", "category"),
      columns = budget_columns,
      detail = custom_detail
    )
  )
  observeEvent(input$tbl_detail, {
    show_detail_modal(input$tbl_detail$id, raw_movements, custom_detail)
  })
}
```
