---
title: "Introduction to treetable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to treetable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(treetable)
```

treetable aggregates a flat data frame into an arbitrary-depth hierarchy with
subtotals at every level ([hierarchical_table()]), and renders that hierarchy
as an interactive, expandable/collapsible tree table built on `htmlwidgets`
([tree_table()]). Everything — row colors, fonts, column widths, number
formats and the per-leaf detail modal — is configured from R, without
touching CSS or JavaScript.

## 1. Aggregating a hierarchy with `hierarchical_table()`

`hierarchical_table()` takes a flat data frame, a vector of hierarchy columns
(from most general to most specific), and a vector of numeric columns to sum
at every level:

```{r}
sales <- tibble::tibble(
  area     = c("Sales", "Sales", "Sales", "Costs", "Costs"),
  category = c("Product A", "Product A", "Product B", "Payroll", "Rent"),
  region   = c("North", "South", "North", "North", "South"),
  amount   = c(1000, 750, 500, -1200, -300)
)

hierarchical_table(sales, levels = c("area", "category"), values = "amount")
```

The result is a long-format tibble: one row per node of the tree (`level_1`,
..., `leaf`), with a subtotal already computed for every group and subgroup.
Add as many levels as you need — `hierarchical_table()` recurses through
however many columns you list in `levels`.

## 2. `tree_table()` with the default theme

`tree_table()` renders the same kind of hierarchy as an interactive tree
table. Unlike `hierarchical_table()`, it expects one or two numeric columns
*per period* you want to compare side by side (e.g. `Q1_budget`/`Q1_actual`),
plus an `id` column identifying each leaf row:

```{r}
budget_vs_actual <- tibble::tibble(
  area = c("Sales", "Sales", "Costs"),
  category = c("Product A", "Product B", "Payroll"),
  id = c("v-a", "v-b", "c-payroll"),
  Q1_budget = c(1000, 500, 300), Q1_actual = c(950, 600, 280),
  Q2_budget = c(1100, 550, 300), Q2_actual = c(1200, 500, 310)
)

tree_table(
  budget_vs_actual,
  levels = c("area", "category"),
  columns = c("Q1", "Q2"),
  suffix_a = "_budget", suffix_b = "_actual",
  label_a = "Budget", label_b = "Actual"
)
```

With two metrics per column, treetable automatically adds a `VAR %` column
showing the percentage variance between them. Pass `suffix_b = NULL` to show
a single metric per column instead, with no variance column.

## 3. Custom theme and column formats

Use [treetable_theme()] to restyle the table's colors, fonts and column
widths, and [column_format()] to control how each metric is displayed
(currency, percentage, decimals, thousands separator):

```{r}
tree_table(
  budget_vs_actual,
  levels = c("area", "category"),
  columns = c("Q1", "Q2"),
  suffix_a = "_budget", suffix_b = "_actual",
  label_a = "Budget", label_b = "Actual",
  theme = treetable_theme(
    header_bg = "#102a43",
    subtotal_bg = "#e3f2fd",
    positive_var_color = "#1a7f37",
    negative_var_color = "#cf222e"
  ),
  format_a = column_format(type = "currency", decimals = 0),
  format_b = column_format(type = "currency", decimals = 0)
)
```

## 4. Per-leaf detail modal in Shiny

Passing a non-`NULL` `detail` to `tree_table()` shows a detail button on
every leaf row. Clicking it sets the Shiny input
`input[[paste0(output_id, "_detail")]]` to `list(id, nonce)`; call
[show_detail_modal()] from an `observeEvent()` to react to it.

### Automatic table

[detail_table()] builds a modal that filters raw, pre-aggregation rows by
leaf id and shows them as a table (using `reactable` when installed, falling
back to `DT`, then to a plain HTML table):

```{r, eval = FALSE}
library(shiny)

raw_movements <- tibble::tibble(
  id = c("v-a", "v-a", "v-b", "c-payroll"),
  date = as.Date(c("2025-01-05", "2025-01-19", "2025-01-11", "2025-01-31")),
  amount = c(600, 400, 500, -300)
)
detail <- detail_table(raw_movements, id_col = "id")

ui <- fluidPage(treetableOutput("tbl"))
server <- function(input, output) {
  output$tbl <- renderTreetable(
    tree_table(
      budget_vs_actual,
      levels = c("area", "category"),
      columns = c("Q1", "Q2"),
      suffix_a = "_budget", suffix_b = "_actual",
      detail = detail
    )
  )
  observeEvent(input$tbl_detail, {
    show_detail_modal(input$tbl_detail$id, raw_movements, detail)
  })
}
shinyApp(ui, server)
```

### Custom UI

Pass a `function(id, raw_data)` instead to render anything you want in the
modal — your own table, a plot, free text:

```{r, eval = FALSE}
custom_detail <- function(id, raw_data) {
  rows <- raw_data[raw_data$id == id, ]
  tagList(
    h4(paste("Movements for", id)),
    renderTable(rows)
  )
}

server <- function(input, output) {
  output$tbl <- renderTreetable(
    tree_table(
      budget_vs_actual,
      levels = c("area", "category"),
      columns = c("Q1", "Q2"),
      suffix_a = "_budget", suffix_b = "_actual",
      detail = custom_detail
    )
  )
  observeEvent(input$tbl_detail, {
    show_detail_modal(input$tbl_detail$id, raw_movements, custom_detail)
  })
}
```
